---
title: "**Practical Machine Learning Course Project**"
author: Aneesh
output:
  html_document:
    highlight: pygments
    theme: cerulean
    keep_md: true
    toc: true
---

# Background

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively.
These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks.
One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants.
They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here:
http://web.archive.org/web/20161224072740/http:/groupware.les.inf.puc-rio.br/har
 (see the section on the Weight Lifting Exercise Dataset).

# Exploratory Data Analysis

```{r}
filename1 <- 'pml-training.csv'
filename2 <- 'pml-testing.csv'

if (!file.exists(filename1) && !file.exists(filename2))
{
	download.file('https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv', filename1, method='curl')
	download.file('https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv', filename2, method='curl')
}

training_full <- read.csv(filename1)
testing <- read.csv(filename2)

dim(training_full)
dim(testing)
str(training_full)
```

# Loading the necessary libraries

```{r}
library(caret)
library(rattle)
```

# Cleaning the data

Removing columns with missing or insufficient data:
```{r}
trainingClean <- training_full[,colSums(is.na(training_full) | training_full == '' | training_full == ' #DIV/0!') == 0]

```

Removing unnecessary variables:
```{r}
trainingClean <- trainingClean[,-c(1:7)] # removing metadata

head(trainingClean, 2)
```

Factoring the `classe` variable:

```{r}
trainingClean$classe <- as.factor(trainingClean$classe)
```

# Partitioning the data

Now, let's divide this training data further into training and validation subsets.

```{r}
set.seed(3314)
inTrain <- createDataPartition(y=trainingClean$classe, p=.75, list=FALSE)
training <- trainingClean[inTrain,]
validation <- trainingClean[-inTrain,]

dim(training)
dim(validation)
```

# Adding parallel processing

As training some of these models is likely to take a lot of time in computation, let's add parallel processing.

```{r}
library(parallel)
library(doParallel)
cluster <- makeCluster(detectCores() - 1) # convention to leave 1 core for OS
registerDoParallel(cluster)
```

# Creating the models

Let's first set up the `trControl` variable for the `train` function.

```{r}
control <- trainControl(method='cv', number=5, verboseIter=F, allowParallel=T)
```

## Linear Discriminant Analysis

```{r, cache=TRUE}
# Model and prediction
mod_lda <- train(classe ~ ., data=training, method='lda')
pred_lda <- predict(mod_lda, validation)
conf_lda <- confusionMatrix(pred_lda, validation$classe)

# Confusion matrix
conf_lda
```

## Decision trees

```{r}
mod_dt <- train(classe ~ ., data=training, method='rpart', trControl=control, tuneLength=6)
fancyRpartPlot(mod_dt$finalModel)

# Prediction
pred_dt <- predict(mod_dt, validation)
conf_dt <- confusionMatrix(pred_dt, validation$classe)

# Confusion Matrix
conf_dt
```

## Random Forests

```{r, cache=TRUE}
mod_rf <- train(classe ~ ., data=training, method='rf', trControl=control, tuneLength=6)

# Prediction
pred_rf <- predict(mod_rf, validation)
conf_rf <- confusionMatrix(pred_rf, validation$classe)

# Confusion Matrix
conf_rf
```

## Models with their accuracy and out of sample error

```{r, cache=TRUE}
models <- c("Random Forest", "Linear Discriminant Analysis", "Decision Trees")
accuracy <- c(conf_rf$overall[1], conf_lda$overall[1], conf_dt$overall[1])
oos_err <- 1 - accuracy # out of sample error

data.frame(Accuracy = accuracy, Out_of_Sample_Error = oos_err, row.names = models)
```
As we can see, the random forest model has the best accuracy or the least out of sample error. Let's use this model to predict the outcomes of the testing dataset.

# Predicting the testing set

```{r}
pred_finalmodel <- predict(mod_rf, testing)

pred_finalmodel
```

# De-register parallel processing cluster

```{r}
stopCluster(cluster)
registerDoSEQ()
```
